---
title: "Getting started with phylorugD1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with phylorugD1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## The biological problem

Phylogenomic studies routinely produce multiple trees from different
methods and datasets. When researchers analyse the same set of taxa
using IQ-TREE, GHOST, and ASTRAL — or using different data types such
as Ultraconserved Elements (UCEs) and partitioned sequence data — they
often obtain slightly different topologies or different levels of node
support. This disagreement between analyses is called **incongruence**
(Steenwyk et al. 2023).

The problem is not that incongruence exists. The problem is communicating
it honestly. Current practice is to publish one tree — usually the
maximum likelihood tree — and move everything else to supplementary
materials. A reader looking at the published figure cannot tell which
nodes are robust across all five analyses and which nodes are contested.
This is a transparency problem.

## The rug plot method

The rug plot method addresses this by drawing a compact colour-coded
grid at each internal node of a backbone tree. Each cell in the grid
represents one alternative analysis, shaded from white (clade absent
or no support) to black (full support). Nodes supported unanimously
by all analyses are shown as a solid dot — maximally clean. Nodes
where analyses disagree show a mixed grid that a reader can interpret
at a glance.

The method was first introduced by Wheeler (1995) under the name
*space plots*, later called *Navajo rugs* (Giribet 2003) and
*topological congruence plots* (Wheeler & Hayashi 1998). Despite thirty
years of use in the literature, no automated R tool existed to produce
these visualizations. **phylorugD1** is the first implementation.

## Who should use this package

Any researcher who:

- runs more than one phylogenomic analysis on the same dataset
- uses IQ-TREE, ASTRAL, GHOST, RAxML, MrBayes, or BEAST
- wants to show node support from multiple analyses in a single figure
- works in R and needs a reproducible, scriptable workflow

The package is organism-agnostic. It was developed and demonstrated on
dung beetles but works on any phylogenomic dataset.

---

## Installation

```{r install}
pak::pak("mdrifathahamed/phylorugD1")
```

---

## The demonstration dataset

This vignette uses a phylogenomic dataset of 289 dung beetle taxa
(*Coleoptera: Scarabaeinae* and *Aphodiinae*) assembled by the Tarasov
Lab at the Finnish Museum of Natural History (LUOMUS), University of
Helsinki. Five analyses were performed:

| Analysis | Method | Data type |
|---|---|---|
| IQ-TREE UCE | Maximum likelihood | UCE |
| IQ-TREE partitioned | Maximum likelihood | Partitioned |
| GHOST partitioned | Heterotachous ML | Partitioned |
| ASTRAL UCE | Coalescent | UCE |
| ASTRAL partitioned | Coalescent | Partitioned |

The IQ-TREE UCE tree is used as the backbone — the most data-rich,
most methodologically standard analysis. The other four are comparison
trees.

---

## Step-by-step pipeline

### Step 1 — Read raw trees

`read_trees_from_dir()` reads all tree files from a folder into a
named list of `phylo` objects. It supports Newick and NEXUS formats.

```{r step1}
library(phylorugD1)
library(readxl)
library(phytools)

raw_trees <- read_trees_from_dir(
  dir    = "data/raw_trees/70p",
  ext    = "tre",
  format = "newick"
)

names(raw_trees)
# [1] "70p_ASTRAL_partition_entropy.tre" "70p_ASTRAL_uce.tre"
# [3] "70p_ghost.tre"                    "70p_partition_entropy.tre"
# [5] "70p_uce.tre"
```

### Step 2 — Root before translating

The outgroup species must be removed before translation because their
tip labels are still in raw specimen code format at this stage.

```{r step2}
rooted_trees <- lapply(raw_trees, function(tr) {
  root(tr,
       outgroup     = c("NicorbUCE", "NicvesUCE"),
       resolve.root = TRUE)
})
```

### Step 3 — Translate tip labels

`translate_tree_tips()` renames tip labels from museum specimen codes
to full species names using a lookup table stored in `biogeo.xlsx`.
The lookup table has two columns: `from` (original code) and `to`
(species name).

```{r step3}
biogeo <- read_excel("data/biogeo.xlsx", sheet = "BioGeo")

translated_trees <- lapply(rooted_trees, function(tr) {
  translate_tree_tips(
    phy      = tr,
    data     = biogeo,
    from_col = "from",
    to_col   = "to"
  )
})

head(translated_trees[[1]]$tip.label)
# [1] "Amphistomus_primonactus_Amppri_UCE"
# [2] "Coptodactyla_brooksi_CopbroUCE"
```

### Step 4 — Extract the ingroup

The outgroup is removed using `phytools::findMRCA()` to locate the
most recent common ancestor of the ingroup, then `extract.clade()` to
keep only the ingroup taxa. This reduces the dataset from 316 to 289
taxa.

```{r step4}
processed_trees <- lapply(translated_trees, function(tr) {
  tr   <- ladderize(tr)
  node <- phytools::findMRCA(
    tr,
    c("Heteronitis_ragazzii_STL10032",
      "Aphodius_immundus_ST003")
  )
  extract.clade(tr, node)
})
names(processed_trees) <- names(raw_trees)

Ntip(processed_trees[[1]])
# [1] 289
```

### Step 5 — Validate taxa

`check_same_taxa()` confirms that all five trees share exactly the
same 289 taxa before any downstream analysis. This must return `TRUE`
before continuing.

```{r step5}
check_same_taxa(processed_trees)
# ✅ All trees contain the same set of taxa.
# [1] TRUE
```

### Step 6 — Define backbone and comparison trees

```{r step6}
backbone  <- processed_trees$`70p_uce.tre`

tree_list <- processed_trees[c(
  "70p_partition_entropy.tre",
  "70p_ghost.tre",
  "70p_ASTRAL_uce.tre",
  "70p_ASTRAL_partition_entropy.tre"
)]
```

### Step 7 — Build the node presence matrix

`node_presence_matrix()` is the core function. For each of the 288
internal nodes in the backbone tree, it searches each comparison tree
for the same clade and records whether it is present.

With `use_support = FALSE`, it returns a binary matrix where `1` means
the clade exists in that analysis and `0` means it does not.

```{r step7}
rug_mt <- node_presence_matrix(
  backbone    = backbone,
  trees       = tree_list,
  use_support = FALSE
)

dim(rug_mt)
# [1] 288   5

head(rug_mt)
#      node_id  70p_partition  70p_ghost  70p_ASTRAL_uce  70p_ASTRAL_part
# [1,]     290            1          1               1               1
# [2,]     291            1          1               1               1
# [3,]     292            1          0               1               1
```

Node 292 is a real biological result. The clade at that node exists in
IQ-TREE partitioned, both ASTRAL analyses, but not in GHOST. This is
topological incongruence — a genuine disagreement between methods
about whether those species form a group.

### Step 8 — Build the grid layout

`rug_layout_map()` assigns each analysis to a fixed position in the
2×2 grid. The layout is the same at every node across the entire tree,
so a reader can learn it once and interpret the whole figure.

```{r step8}
layout_df <- rug_layout_map(rug_mt, n_rows = 2, n_cols = 2)

nrow(layout_df)
# [1] 1152  (288 nodes x 4 analyses)
```

The grid positions are:

```
┌───────────────────┬──────────────────────┐
│ IQ-TREE part      │  GHOST part          │
├───────────────────┼──────────────────────┤
│ ASTRAL UCE        │  ASTRAL part         │
└───────────────────┴──────────────────────┘
```

### Step 9 — RF similarity plot

`rf_similarity_plot()` computes the Robinson-Foulds distance between
every pair of trees and projects the result into 2D using
Multidimensional Scaling (MDS). Trees close together in the plot have
similar topologies. Trees far apart disagree substantially.

```{r step9}
all_trees        <- processed_trees
names(all_trees) <- c("ASTRAL_part", "ASTRAL_uce",
                      "GHOST", "IQTREE_part", "IQTREE_uce")
class(all_trees) <- "multiPhylo"

rf_similarity_plot(all_trees)
```

### Step 10 — Draw the rug plot

`plot_node_rug()` must be called after `plot.phylo()` — it reads node
coordinates from the active plot environment and draws the rug grid
at each uncertain node.

```{r step10}
# Colour palette: white = absent, black = full support
pal_info        <- list()
pal_info$pal    <- gray.colors(64, start = 0.95, end = 0.00)
pal_info$breaks <- seq(0, 1, length.out = 64)

map_to_color <- function(val, pal_info) {
  if (val == 0) return("white")
  pal_info$pal[cut(val, breaks = pal_info$breaks,
                   include.lowest = TRUE)]
}

# Split nodes
rug_mt_variable <- rug_mt[
  !apply(
    rug_mt[, -1, drop = FALSE],
    1,
    function(x) all(x == 1)
  ), , drop = FALSE
]

rug_mt_unanimous <- rug_mt[
  apply(
    rug_mt[, -1, drop = FALSE],
    1,
    function(x) all(x == 1)
  ), , drop = FALSE
]
# Plot backbone tree
plot.phylo(backbone, show.tip.label = TRUE, cex = 0.4,
           label.offset = 0.001, no.margin = TRUE, edge.width = 1.5)

# Solid dots at unanimous nodes
last_pp  <- get("last_plot.phylo", envir = ape::.PlotPhyloEnv)
node_ids <- rug_mt_unanimous[, 1]
points(last_pp$xx[node_ids], last_pp$yy[node_ids],
       pch = 16, cex = 0.6, col = "black")

# Compute cell size
dy         <- median(diff(sort(last_pp$yy[1:Ntip(backbone)])))
x_per_inch <- diff(last_pp$x.lim) / par("pin")[1]
y_per_inch <- diff(last_pp$y.lim) / par("pin")[2]
cell_h     <- dy * 0.18
cell_w     <- cell_h * (x_per_inch / y_per_inch)

# Draw rug grid
plot_node_rug(
  tree         = backbone,
  rug_mt       = rug_mt_variable,
  cell_h       = cell_h,
  cell_w       = cell_w,
  x_offset     = 0.04,
  y_offset     = 0,
  map_to_color = map_to_color,
  pal_info     = pal_info
)
```

---

## Reading the output

**Solid black dot** — all four comparison analyses support this node
with maximum support. The node is robust. No further investigation
needed.

**Coloured grid** — at least one analysis differs. Read the grid:

```
┌──────────────┬──────────────┐
│ IQ-TREE part │  GHOST part  │
├──────────────┼──────────────┤
│  ASTRAL UCE  │ ASTRAL part  │
└──────────────┴──────────────┘
```

- **Black cell** — clade present and fully supported in this analysis
- **White cell** — clade absent from this analysis entirely
- **Grey cell** — clade present but with partial support (when
  `use_support = TRUE`)

A node where IQ-TREE cells are black but ASTRAL cells are white means
the maximum likelihood and coalescent methods disagree. This is
biologically meaningful — it may indicate incomplete lineage sorting
or rapid diversification at that node.

---

## Tuning the figure

Two parameters control the visual appearance:

```{r tuning}
cell_h <- dy * 0.18   # increase for bigger cells, decrease for smaller
x_offset <- 0.04      # increase to move right, decrease to move left
```

For large trees (200+ taxa) we recommend saving to PDF with
`width = 16` and `height = 40` to give each node enough space.

---

## References

Giribet, G. (2003). Stability in phylogenetic formulations and its
relationship to nodal support. *Systematic Biology*, 52(4), 554–564.

Steenwyk, J. L., Li, Y., Zhou, X., Shen, X. X., & Rokas, A. (2023).
Incongruence in the phylogenomics era. *Nature Reviews Genetics*,
24(12), 834--850.

Wheeler, W. C. (1995). Sequence alignment, parameter sensitivity, and
the phylogenetic analysis of molecular data. *Systematic Biology*,
44(3), 321–331.

Wheeler, W. C., & Hayashi, C. Y. (1998). The phylogeny of the extant
chelicerate orders. *Cladistics*, 14, 173–192.
